Powering Hidden Markov Model by Generative Models

### 2019-08-12 manuscript/main.tex
structure orginized

\usepackage{caption}
\usepackage{bm}
\newcommand{\ubar}[1]{\mkern2mu\underline{\mkern-2mu #1\mkern-2mu}\mkern2mu}
% \allowdisplaybreaks
\usepackage{mystyle}
\newcommand{\ubm}[1]{\ubar{\bm{#1}}}
\newcommand{\ubmr}[2]{\ubar{\bm{#1}}^{(#2)}}

@@

\newcommand{\bmtr}[3]{\bm{#1}^{(#3)}_{#2}}
\newcommand{\smtr}[3]{{#1}^{(#3)}_{#2}}

@@

\section{Preliminaries}
\begin{figure}[!h]
  \centering
  \begin{tikzpicture}
    \tikzstyle{enode} = [thick, draw=black, ellipse, inner sep = 1pt,  align=center]
    \tikzstyle{nnode} = [thick, rectangle, rounded corners = 2pt,minimum size = 0.8cm,draw,inner sep = 2pt]
    \node[enode] (g1) at (-0.5,1.8) {$p(\bm{x}| s=1; \bm{\Phi}_{1})$};
    \node[enode] (g2) at (-0.5,0.5) {$p(\bm{x}| s=2; \bm{\Phi}_{2})$};
    \node[enode] (gs) at (-0.5, -1.8) {$p(\bm{x}| s=|\Ss|; \bm{\Phi}_{|\Ss|})$};
    \node[enode] (x) at (4.5,1.5){$\ubm{x}\sim p(\ubm{x};\bm{H})$};

@@

In the framework of HMM, at each time instance, signal $\bm{x}_t$ is generated by density $p(\bm{x}_t| s_t; \bm{\Phi}_{s_t})$, and $s_t$ is decided by the hidden markov process. This process gives us the probabilistic model $p(\ubm{x};\bm{H})$
\section{Generator-mixed HMM}

@@

    \draw[->] (zK) -- (gs);
    \draw[->] (gs) -- (1.8, -1.8);
  \end{tikzpicture}
  \caption{Source $s$ powered by generator mixed generative model (GenHMM).}
  \vspace{0.1cm}
\end{figure}

In this section, we introduce a neural network based generator-mixed HMM model. We term it as GenHMM. GenHMM has a hidden markov process as typical HMM, while the probabilistic model of GenHMM for each hidden state is induced by mixture of neural network based generators. For $s\in \Ss$, we define the state's probabilistic model in GenHMM as
\begin{equation}
  p(\bm{x}| s; \bm{\Phi}_{s}) = \sum_{\kappa=1}^{K}\pi_{s, \kappa} p(\bm{x}| s, \kappa; \bm{\Phi}_{s, \kappa})
\end{equation}
where $\kappa$ is a random variable following a categorical distribution, and $\pi_{s, \kappa} = p(\kappa | \bm{x}, s; \bm{\Phi}_{s, \kappa})$. Naturally $\sum_{\kappa = 1}^{K} \pi_{s, \kappa}= 1$. Denote $\bm{\pi}_{s} = [\pi_{s,1}, \pi_{s,2}, \cdots, \pi_{s,K}]$,
$\bm{\Phi}_s = \left\{ \pi_{s, \kappa}, \bm{\Phi}_{s, \kappa}| \kappa = 1, 2, \cdots, K \right\}$,

\subsubsection{Generator Mixed HMM (GenHMM)}\label{subsubsec:GenHMM}
In this subsection, we specify the GenHMM model and give the learning algorithm of this model.
Base on the generative model assumption, for each state $s$, the probabilistic generative model is

\subsubsection{Flow structure}

\section{Learning Algorithm of GenHMM}
\subsection{Problem Statement}
Given a empirical distribution $\hat{p}(\ubm{x}) = \frac{1}{R}\sum_{r=1}^{R} \delta_{\ubmr{x}{r}}(\ubm{x})$. We want to find a probabilistic model such that:
\begin{equation}
  \min KL(\hat{p}(\ubm{x})\| p(\ubm{x}))
\end{equation}
where $KL(\cdot\|\cdot)$ denotes the Kullback-Leibler divergence.

When we use HMM to model the empirical distribution and approach the unknown true distribution, the problem boils down to:
\begin{equation}
  \uargmax{\bm{H} \in \Hh} p(\ubm{X}; \bm{H})
\end{equation}
where $\ubm{X} = \left[ \ubmr{x}{1}, \ubmr{x}{2}, \cdots, \ubmr{x}{R} \right]$

The problem can be reformulated as
\begin{equation}\label{eq:ml-of-hmm}
  \uargmax{\bm{H} \in \Hh} \sum_{r=1}^{R}\log\,p(\ubmr{x}{r}; \bm{H})
\end{equation}
for independent identical distributed assumption of $\ubm{x}$.

\subsection{Proposal}

Since model $\bm{H}$ contains hidden sequential variable $\ubm{s}$, we can not directly solve the maximum likelihood problem in \ref{eq:ml-of-hmm}. We use expectation maximization (EM) to address the hidden variable problem by
\begin{itemize}
\item E-step: % the posterior probability of $\ubm{s}$:
  % \begin{equation}
  %   p(\ubm{s}|\ubm{x})
  % \end{equation}
  The ``expected likelihood'' function:
  \begin{equation}\label{eq:em-q-funciton}
    \Qq(\bm{H}; \bm{H}^{\mathrm{old}}) = \ \sum_{r=1}^{R}E_{p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})}\left[\log\,p(\ubmr{x}{r}, \ubmr{s}{r}; \bm{H}) \right]
  \end{equation}
\item M-step: the optimization step:
  \begin{equation}\label{eq:em-m-opt}
    \umax{\bm{H}} \Qq(\bm{H}; \bm{H}^{\mathrm{old}})
  \end{equation}
\end{itemize}

The \eqref{eq:em-m-opt} can be reformulated as:
\begin{equation}\label{eq:m-step-subs}
  \umax{\bm{H}} \Qq(\bm{H}; \bm{H}^{\mathrm{old}}) = \umax{\bm{q}}\Qq(\bm{q}; \bm{H}^{\mathrm{old}}) + \umax{{A}}\Qq({A}; \bm{H}^{\mathrm{old}}) + \umax{\bm{\Phi}}\Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}})
\end{equation}
where
\begin{align}
  \Qq(\bm{q}; \bm{H}^{\mathrm{old}}) &= \sum_{r=1}^{R} \EE_{p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \left[\log\,p(\smtr{s}{1}{r}; \bm{q})\right] \label{eq:init-distribution-update}\\
  \Qq(\bm{A}; \bm{H}^{\mathrm{old}}) &= \sum_{r=1}^{R} \EE_{p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \left[\log\,\sum_{t=1}^{T^{(r)}-1}p(\smtr{s}{t+1}{r}|\smtr{s}{t}{r}; {A})\right] \label{eq:transition-update}\\
  \Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}}) &= \sum_{r=1}^{R} \EE_{p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \left[\log\,p(\ubmr{x}{r} | \ubmr{s}{r}; \bm{\Phi})\right]\label{eq:generative-model-update}
\end{align}

@@

>>>>

\usepackage{caption}
\usepackage{bm}
\newcommand{\ubar}[1]{\mkern2mu\underline{\mkern-2mu #1\mkern-2mu}\mkern2mu}
% \allowdisplaybreaks
\usepackage{mystyle}
\newcommand{\ubm}[1]{\ubar{\bm{#1}}}
\newcommand{\ubmr}[2]{\ubar{\bm{#1}}^{#2}}

@@

\newcommand{\bmtr}[3]{\bm{#1}^{#3}_{#2}}
\newcommand{\smtr}[3]{{#1}^{#3}_{#2}}

@@

\section{Generator-mixed HMM}

\begin{figure}[!h]
  \centering
  \begin{tikzpicture}
    \tikzstyle{enode} = [thick, draw=black, ellipse, inner sep = 1pt,  align=center]
    \tikzstyle{nnode} = [thick, rectangle, rounded corners = 2pt,minimum size = 0.8cm,draw,inner sep = 2pt]
    \node[enode] (g1) at (-0.5,1.8) {$p(\bm{x}| s=1; \bm{\Phi}_{1})$};
    \node[enode] (g2) at (-0.5,0.5) {$p(\bm{x}| s=2; \bm{\Phi}_{2})$};
    \node[enode] (gs) at (-0.5, -1.8) {$p(\bm{x}| s=|\Ss|; \bm{\Phi}_{|\Ss|})$};
    \node[enode] (x) at (4.5,1.5){$\ubm{x}\sim p(\ubm{x};\bm{H})$};

@@

In the framework of HMM, at each time instance, signal $\bm{x}_t$ is generated by density $p(\bm{x}_t| s_t; \bm{\Phi}_{s_t})$, and $s_t$ is decided by the hidden markov process. This process gives us the probabilistic model $p(\ubm{x};\bm{H})$
\subsection{Emission Model of GenHMM}

@@

    \draw[->] (zK) -- (gs);
    \draw[->] (gs) -- (1.8, -1.8);
  \end{tikzpicture}
  \caption{Source of state $s$ in GenHMM.}
  \label{fig:gen-mix}
  \vspace{0.1cm}
\end{figure}

In this section, we introduce a neural network based generator-mixed HMM model. We term it as GenHMM. GenHMM has a hidden markov process as typical HMM, while the probabilistic model of GenHMM for each hidden state is induced by mixture of neural network based generators. For given state $s\in \Ss$, we define its probabilistic model in GenHMM as
\begin{equation}
  p(\bm{x}| s; \bm{\Phi}_{s}) = \sum_{\kappa=1}^{K}\pi_{s, \kappa} p(\bm{x}| s, \kappa; \bm{\theta}_{s, \kappa}),
\end{equation}
where $\kappa$ is a random variable following a categorical distribution, and $\pi_{s, \kappa} = p(\kappa | s; \bm{\Phi}_{s, \kappa})$. Naturally $\sum_{\kappa = 1}^{K} \pi_{s, \kappa}= 1$. Denote $\bm{\pi}_{s} = [\pi_{s,1}, \pi_{s,2}, \cdots, \pi_{s,K}]$,
$\bm{\theta}_s = \left\{ \bm{\theta}_{s, \kappa}| \kappa = 1, 2, \cdots, K \right\}$. 
We define $p(\bm{x}| s, \kappa; \bm{\theta}_{s, \kappa})$ as induced distribution by a generator $\bm{g}_{s,\kappa}: \RR^{N}\rightarrow\RR^{N}$, such that $\bm{x}=\bm{g}_{s, \kappa}(\bm{z})$, where $\bm{z}\sim p_{s,\kappa}(\bm{z})$. By change of variable, we have
\begin{equation}\label{eq:changel-variable}
  p(\bm{x}| s, \kappa; \bm{\theta}_{s, \kappa}) = p_{s,\kappa}(\bm{z})\bigg| \det\left( \pd{\bm{g}_{s,\kappa}(\bm{z})}{\bm{z}} \right)\bigg|^{-1}.
\end{equation}

For a state $s$ of GenHMM, the induced probability distribution is illustrated in Figure~\ref{fig:gen-mix}.

\subsection{Problem Statement}

For a given dataset, we denote its empirical distribution by $\hat{p}(\ubm{x}) = \frac{1}{R}\sum_{r=1}^{R} \delta_{\ubmr{x}{r}}(\ubm{x})$, where superscipt $(\cdot)^{r}$ denotes the index of sequential signal. The natural question is how to use GenHMM to represent the dataset. Alternative, we are looking for the answer to question:
\begin{equation}
  \umin{\bm{H}\in \Hh} KL(\hat{p}(\ubm{x})\| p(\ubm{x};\bm{H}))
\end{equation}
where $KL(\cdot\|\cdot)$ denotes the Kullback-Leibler divergence. The KL divergence problem can be boiled down to a maximizing a loglikelihood problem:
\begin{equation}\label{eq:ml-of-hmm}
  \uargmax{\bm{H} \in \Hh} \sum_{r=1}^{R}\log\,p(\ubmr{x}{r}; \bm{H}).
\end{equation}

With such a model GenHMM at hand, the remaining problems are:
\begin{itemize}
\item How to realize GenHMM by neural networks to have high modeling capacity?
\item How to train GenHMM to sovle problem in \eqref{eq:ml-of-hmm} using practical algorithm?
\item Would the training of GenHMM converges?
\end{itemize}
We would answer these questions in the following section.

\section{Practical Solution for GenHMM}

In this section, we provide answers to the questions at the end of last section.

Since model $\bm{H}$ contains hidden sequential variable $\ubm{s}$ and $\ubm{\kappa}$, we can not directly solve the maximum likelihood problem in \ref{eq:ml-of-hmm}. We use expectation maximization (EM) to address the hidden variable problem by
\begin{itemize}
\item E-step: % the posterior probability of $\ubm{s}$:
  % \begin{equation}
  %   p(\ubm{s}|\ubm{x})
  % \end{equation}
  The ``expected likelihood'' function:
  \begin{equation}\label{eq:em-q-funciton}
    \Qq(\bm{H}; \bm{H}^{\mathrm{old}}) = \EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})}\left[ \log\,p(\ubm{x}, \ubm{s}, \ubm{\kappa}; \bm{H})\right],
  \end{equation}
  where $\EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})}\left[ \cdot\right]$ denotes the expectation operator by distribution $\hat{p}(\ubm{x})$ and $p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})$.
\item M-step: the optimization step:
  \begin{equation}\label{eq:em-m-opt}
    \umax{\bm{H}} \Qq(\bm{H}; \bm{H}^{\mathrm{old}})
  \end{equation}
\end{itemize}

The \eqref{eq:em-m-opt} can be reformulated as:
\begin{align}\label{eq:m-step-subs}
  &\umax{\bm{H}} \Qq(\bm{H}; \bm{H}^{\mathrm{old}}) \nonumber \\
  =&\umax{\bm{q}}\Qq(\bm{q}; \bm{H}^{\mathrm{old}}) + \umax{{A}}\Qq({A}; \bm{H}^{\mathrm{old}}) 
  + \umax{\bm{\Phi}}\Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}})
\end{align}
where
\begin{align}
  \Qq(\bm{q}; \bm{H}^{\mathrm{old}}) 
    &=\EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})} \left[ \log\,p({s}_{1})  \right] \nonumber\\
    &= \EE_{\hat{p}(\ubm{x}),p(\ubm{s}| \ubm{x}; \bm{H}^{\mathrm{old}})} \left[ \log\,p({s}_{1})  \right]\label{eq:init-distribution-update}\\
  \Qq(\bm{A}; \bm{H}^{\mathrm{old}}) &=\EE_{\hat{p}(\ubm{x}),p(\ubm{s}| \ubm{x}; \bm{H}^{\mathrm{old}})}\left[ \sum_{t}\log\,p({s}_{t+1}|{s}_{t}; \bm{H}) \right] \label{eq:transition-update}\\
  \Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}}) &= \EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})} \left[ \log\,p(\ubm{x}, \ubm{\kappa}| \ubm{s}; \bm{H}) \right]\label{eq:generative-model-update}
\end{align}

@@

\subsection{Modeling $\bm{g}_{s,\kappa}$ by a flow model }
We use feed-forward neural network to implement every generator $\bm{g}_{s,\kappa}$. For notation simplicity, denote $\bm{g}_{s,\kappa}=\bm{g}_{s,\kappa}$. Define  a feed-forward neural network $\bm{g}_{s,\kappa}$ that has multiple hidden layers
$\bm{g}_{s,\kappa}=\bm{g}_{s,\kappa}^{[L]}\circ \bm{g}_{s,\kappa}^{[L-1]}\circ \cdots
\circ \bm{g}_{s,\kappa}^{[1]}$ and is invertible $\bm{f}_{s,\kappa}=\bm{g}_{s,\kappa}^{-1}$, where $\circ$ denotes mapping concatenation. Then the signal flow can be depicted as
\begin{equation*}
  \vspace{-8pt}
  \centering
  \begin{tikzpicture}
    \node (z) at (0,0) {};
    \node at ($(z)-(0.5,0)$){$\bm{z}=\bm{h}_0$};
    \node (xi1) at (1.5,0) {$\bm{h}_1$};
    \node (xi2) at (3,0) {};
    \node (xi3) at (4.5,0){};
    \node (x) at (6,0) {};
    \node at ($(x)+(0.5,0)$){${\bm{x}} = \bm{h}_L$};
    \draw[->] ($(z) + (0.3,0.1)$) -- node[above]{$\bm{g}_{s,\kappa}^{[1]}$} ($(xi1)+(-0.3,0.1)$); 
    \draw[->] ($(xi1)-(0.3,0.1)$) -- node[below]{${\bm{f}}_{s,\kappa}^{[1]}$}($(z) - (-0.3,0.1)$);
    \draw[->] ($(xi1) + (0.3,0.1)$) -- node[above]{$\bm{g}_{s,\kappa}^{[2]}$} ($(xi2)+(-0.3,0.1)$); 
    \draw[->] ($(xi2)-(0.3,0.1)$) -- node[below]{${\bm{f}}_{s,\kappa}^{[2]}$}($(xi1) - (-0.3,0.1)$);
    \draw[->] ($(xi3) + (0.3,0.1)$) -- node[above]{$\bm{g}_{s,\kappa}^{[L]}$} ($(x)+(-0.3,0.1)$); 
    \draw[->] ($(x)-(0.3,0.1)$) -- node[below]{${\bm{f}}_{s,\kappa}^{[L]}$}($(xi3) - (-0.3,0.1)$);
    \draw[dotted,line width = 0.3 mm] (xi2) -- (xi3);
  \end{tikzpicture},
\end{equation*}
where $\bm{g}_{s,\kappa}^{[l]}$ and $\bm{f}_{s,\kappa}^{[l]}$ are the $l$-th layer of $\bm{g}_{s,\kappa}$ and $\bm{f}_{s,\kappa}$, respectively. In a feed-forward neural network, if every layer is invertible,
the full feed-forward neural network is invertible. The inverse
function is given by $\bm{z}=\bm{f}_{s,\kappa}(\bm{x})$. Flow-based
network, proposed in \cite{DBLP:journals/corr/DinhKB14}, is such a
feed-forward neural network, which is further improved in subsequent works \cite{2016arXiv160508803D, 2018arXiv180703039K}. It also has additional advantages as efficient Jacobian computation and low computational complexity. %Computation of inverse mapping for flow-based network and determinant of Jacobin were addressed in the literature.

For a flow-based neural network architecture, let us assume that
the feature $\bm{h}_l$ at the $l$'th layer has two subparts as
$\bm{h}_l = [\bm{h}_{l,a}^{T} \, , \, \bm{h}_{l,b}^{T}]^{T}$ where
$(\cdot)^{T}$ denotes transpose operation. Then considering $\bm{h}_0 = \bm{z}$, we have the following forward and inverse relations between $(l-1)$'th and $l$'th layers:
\begin{equation}\label{eq-gl}
  \begin{array}{l}
    \bm{h}_{l-1} =
    \begin{bmatrix}
      \bm{h}_{l-1,a}\\
      \bm{h}_{l-1,b}
    \end{bmatrix}
    =
    \begin{bmatrix}
      \bm{h}_{l,a}\\
      \bm{m}_a(\bm{h}_{l,a})\odot \bm{h}_{l,b} + \bm{m}_b(\bm{h}_{l,a})
    \end{bmatrix},\vspace{10pt}\\
    \bm{h}_{l} =
    \begin{bmatrix}
      \bm{h}_{l,a}\\
      \bm{h}_{l,b}
    \end{bmatrix}
    =
    \begin{bmatrix}
      \bm{h}_{l-1,a}\\
      \left(  \bm{h}_{l-1,b} - \bm{m}_b(\bm{h}_{l-1,a}) \right)\oslash \bm{m}_a(\bm{h}_{l-1,a}) 
    \end{bmatrix}, \\
  \end{array}  
\end{equation}
where $\odot$ denotes element-wise product, $\oslash$ denotes
element-wise division, and $\bm{m}_a(\cdot), \bm{m}_b(\cdot)$ can be
complex non-linear mappings (implemented by neural networks).
For the flow-based neural network, the determinant of Jacobian matrix is
\begin{equation}\label{eq:cat-jacobian}
  \begin{array}{rl}
    \mathrm{det}(\nabla{\bm{f}_{s,\kappa}}) & = \prod_{l=1}^L \det (\nabla{\bm{f}_{s,\kappa}^{[l]}}),
  \end{array}
\end{equation}
where $\nabla{f_{s,\kappa}^{[l]}}$ is the Jacobian of the transformation from the $l$-th layer to the $(l-1)$-th layer, i.e., the inverse transformation. We compute the determinate of the Jacobian matrix as
\begin{align}\label{eq-hl-determinate}
  \det (\nabla{f_{s,\kappa}^{[l]}})& = \det \left[  \pd{\bm{h}_{l-1}}{\bm{h}_l} \right] \nonumber\\
                               & = \det
                                 \begin{bmatrix}
                                   \bm{I}_a & \mathbf{0} \nonumber\\
                                   \pd{\bm{h}_{l-1,b}}{\bm{h}_{l,a}} & \mathrm{diag}(\bm{m}_a(\bm{h}_{l,a}))
                                 \end{bmatrix}\nonumber\\
                               &= \det \left( \mathrm{diag}(\bm{m}_a(\bm{h}_{l,a})) \right),
\end{align}
where $\bm{I}_a$ is identity matrix and $\mathrm{diag}(\cdot)$ returns a square matrix with the elements of $(\cdot)$ on the main diagnal.
\textcolor{red}{do I still need this?}
Then the pdf is
\begin{align}
  p(\bm{x}) & =  p(\bm{z}) \big| \mathrm{det}(\bm{J}) |_{\bm{z}={\bm{f}}({\bm{x}})}\big| \nonumber\\
            &  = p(\bm{z}) \prod_{l=1}^L \abs{\det\left( \mathrm{diag}(\bm{m}_a(\bm{h}_{l,a}))]  \right)}.
\end{align}
[] describes a \textit{coupling} mapping between layers. Since the coupling has a partial identity mapping, direct concatenation of multiple such coupling mappings would result in a partial identity mapping of the whole neural network $\tilde{\bm{g}}$. Alternating the positions of identity mapping \cite{2016arXiv160508803D} or using $1\times1$ convolution operations \cite{2018arXiv180703039K} before each coupling mapping is used to treat the issue.
              %               \textcolor{blue}{Some reading problem in this sentence: Since the coupling has a partial identity mapping, alternating the positions of identity mapping \cite{2016arXiv160508803D} or using $1\times1$ convolution operations \cite{2018arXiv180703039K} when
              %               concatenating multiple layers together were proposed to avoid the situation that a part of the output is equal to a part of the input.} %In addition, applying affine mapping does not affect its advantage of trivial inverse and determinate computation.
Furthermore, \cite{2016arXiv160508803D}\cite{2018arXiv180703039K} split some hidden
layer signal $\bm{h}$ and model a part of it directly as standard Gaussian to reduce computation and memory burden.
% % \textcolor{blue}{Therefore the dimension
              %               of latent signal $\bm{z}$ to model can be smaller than $N$.}

\subsection{Learning of GenHMM}

\subsubsection{Generative Model Update}
We use notation $\bm{\Pi} = \left\{  \bm{\pi}_{s}| s\in \Ss \right\}$, $\bm{\Theta}=\left\{ \bm{\theta}_s| s\in \Ss \right\}$.
Then the third subproblem, \eqref{eq:m-step-subs}, becomes:
\begin{align}\label{eq:sub-gm}
  &\umax{\bm{\Phi}} \Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}}) = \umax{\bm{\Pi}} \Qq(\bm{\Pi}; \bm{H}^{\mathrm{old}}) + \umax{\bm{\Theta}} \Qq(\bm{\Theta}; \bm{H}^{\mathrm{old}}),
\end{align}
where
\begin{align}
  &\Qq(\bm{\Pi}; \bm{H}^{\mathrm{old}})  =\EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})}\left[  \log\,p(\ubm{\kappa}| \ubm{s}; \bm{H})\right] \nonumber \\
  &\Qq(\bm{\Theta}; \bm{H}^{\mathrm{old}}) =\EE_{\hat{p}(\ubm{x}),p(\ubm{s},\ubm{\kappa}| \ubm{x}; \bm{H}^{\mathrm{old}})}\left[  \log\,p(\ubm{x}| \ubm{s},\ubm{\kappa}; \bm{H})\right] 
\end{align}

The neural network loss can be thus formulated as
\begin{align}\label{eq:obj-q-gen-mix}
  &\Qq(\bm{\Theta}; \bm{H}^{\mathrm{old}}) \nonumber \\
  = &\frac{1}{R}\sum_{r=1}^{R}\sum_{\ubmr{s}{r}}\sum_{\ubmr{\kappa}{r}}{p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \sum_{t=1}^{{T}^{r}}\log\,p(\bm{x}^{(r)} | \smtr{s}{t}{r}, \smtr{\kappa}{t}{r}; \bm{H}) \nonumber \\
  =& \frac{1}{R}\sum_{r=1}^{R} \sum_{t=1}^{{T}^{r}-1} \sum_{\smtr{s}{t}{r}=1}^{|\Ss|}  \sum_{\smtr{\kappa}{t}{r}=1}^{K}p(\smtr{s}{t}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})p(\smtr{\kappa}{t}{r}|\smtr{s}{t}{r}, \ubmr{x}{r}; \bm{H}^{\mathrm{old}}) \nonumber\\
&  \log\, p(\bmtr{x}{t}{r} | \smtr{s}{t}{r}, \smtr{\kappa}{t}{r}; \bm{H}) 
\end{align}

Here, $p(s_t| \ubm{x}, \bm{H}^{\mathrm{old}})$ is computed by forward/backward algorithm. The posterior of $\kappa$ is:
\begin{align}\label{eq:kappa-posterior}
  p(\kappa| s, \ubm{x}; \bm{H}^{\mathrm{old}})
  &=  \frac{p(\kappa, \ubm{x}| s; \bm{H}^{\mathrm{old}})}{p(\ubm{x}| s,\bm{H}^{\mathrm{old}})} \nonumber \\
  & = \frac{\pi_{s, \kappa}^{\mathrm{old}} p(\bm{x}| s, \kappa, \bm{H}^{\mathrm{old}})}{\sum_{\kappa=1}^{K}  \pi_{s, \kappa}^{\mathrm{old}} p(\bm{x}| s, \kappa,\bm{H}^{\mathrm{old}})} 
\end{align}
where the last equation is due to the fact that only $\bm{x}_t$ among sequence $\ubm{x}$ depends on $s_t, \kappa_t$.

\textcolor{red}{expand to neural network here}By substituting \eqref{eq:changel-variable} and \eqref{eq:cat-jacobian} into \eqref{eq:obj-q-gen-mix}, we have loss function for neural networks as
\begin{align}\label{eq:obj-q-gen-mix}
  &\Qq(\bm{\Theta}; \bm{H}^{\mathrm{old}}) \nonumber \\
  =& \frac{1}{R}\hspace{-3pt}\sum_{r=1}^{R}\hspace{-3pt} \sum_{t=1}^{{T}^{r}-1}\hspace{-3pt} \sum_{\smtr{s}{t}{r}=1}^{|\Ss|} \hspace{-3pt} \sum_{\smtr{\kappa}{t}{r}=1}^{K}p(\smtr{s}{t}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})p(\smtr{\kappa}{t}{r}|\smtr{s}{t}{r}, \ubmr{x}{r}; \bm{H}^{\mathrm{old}}) \nonumber\\
&\left[ \log\, p_{\smtr{s}{t}{r}, \smtr{\kappa}{t}{r}}(\bm{f}_{\smtr{s}{t}{r}, \smtr{\kappa}{t}{r}}(\bmtr{x}{t}{r})) + \sum_{l=1}^{L}\log\,| \det (\nabla{\bm{f}_{s,\kappa}^{[l]}})|\right]
\end{align}

In implementation, the generators of GenHMM uses standard Gaussian distribution for variables $\bm{z} = \bm{f}_{s,\kappa}(\bm{x})$, i.e. $p_{s,\kappa}(\bm{z})$ used standard Gaussian density function.

------------------------------

To solve problem \eqref{opm:pi}, we formulate its Lagrange function as
\begin{equation}
  \Ll = \Qq(\bm{\Phi}; \bm{H}^{\mathrm{old}}) + \sum_{\lambda_s} \lambda_s\left( 1-  \sum_{\kappa=1}^{K}\pi_{s, \kappa}  \right).
\end{equation}
Solving $\pd{\Ll}{\pi_{s, \kappa}} = 0$ gives
\begin{equation}
  \pi_{s,\kappa} = \frac{1}{\lambda_s}\sum_{r=1}^{R} \sum_{t=1}^{{T}^{(r)}} p(\smtr{s}{t}{r}=s, \smtr{\kappa}{t}{r} =\kappa| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})
\end{equation}
With condition $\sum_{\kappa=1}^{K} \pi_{s, \kappa}= 1, \forall s = 1, 2, \cdots, |\Ss|$, we have
\begin{equation}
  \lambda_s = \sum_{\kappa=1}^{K}\sum_{r=1}^{R} \sum_{t=1}^{{T}^{(r)}} p(\smtr{s}{t}{r}=s, \smtr{\kappa}{t}{r} =\kappa | \ubmr{x}{r}; \bm{H}^{\mathrm{old}})
\end{equation}
Then the solution to \eqref{opm:pi} is
\begin{equation}\label{eq:mix-latent-parameter-solution}
  \pi_{s, \kappa} = \frac{\sum_{r=1}^{R} \sum_{t=1}^{{T}^{r}} p(\smtr{s}{t}{r} =s, \smtr{\kappa}{t}{r}=\kappa | \ubmr{x}{r}; \bm{H}^{\mathrm{old}}) }{\sum_{k =1}^{K}\sum_{r=1}^{R} \sum_{t=1}^{{T}^{r}-1} p(\smtr{s}{t}{r} =s, \smtr{\kappa}{t}{r}=k | \ubmr{x}{r}; \bm{H}^{\mathrm{old}}) }
\end{equation}
where $p(s, \kappa | \ubm{x}; \bm{H}^{\mathrm{old}}) = p(s| \ubm{x}; \bm{H}^{\mathrm{old}}) p(\kappa | s, \ubm{x}; \bm{H}^{\mathrm{old}})$. $p(s| \ubm{x}; \bm{H}^{\mathrm{old}})$ that can be computed by results of forward/backward in Viterbi algorithm, while $p(\kappa | s, \ubm{x}; \bm{H}^{\mathrm{old}})$. $p(s| \ubm{x}; \bm{H}^{\mathrm{old}})$ is given by \eqref{eq:kappa-posterior}.

-----------------------

\subsubsection{Initial Probability Update}
\eqref{eq:init-distribution-update} can be written as:
\begin{align}
  &\Qq(\bm{q}; \bm{H}^{\mathrm{old}}) \nonumber\\
  &=\frac{1}{R} \sum_{r=1}^{R}\sum_{\ubmr{s}{r}} {p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \log\,p(\smtr{s}{1}{r}) \nonumber \\
% & = \frac{1}{R}\sum_{r=1}^{R}\sum_{\smtr{s}{1}{r}=1}^{|\Ss|}\sum_{\smtr{s}{2}{r}=1}^{|\Ss|}\cdots \sum_{\smtr{s}{T^{r}}{r}}^{{|\Ss|}} {p(\smtr{s}{1}{r}, \smtr{s}{2}{r}, \cdots, \smtr{s}{T^{r}}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \log\,p(\smtr{s}{1}{r}) \\
& = \frac{1}{R}\sum_{r=1}^{R}\sum_{\smtr{s}{1}{r}=1}^{|\Ss|}{p(\smtr{s}{1}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \log\,p(\smtr{s}{1}{r}) 
\end{align}

Since $p(\smtr{s}{1}{r})$ is the probability of initial state of GenHMM $\bm{H}$ for $r$-th sequential sample, actually $q_i = p({s}_{1} =i) $, $i= 1, 2, \cdots, |\Ss|$ for $\bm{H}$. Solution to problem:
\begin{align}
  \bm{q} &= \uargmax{\bm{q}} \Qq(\bm{q}; \bm{H}^{\mathrm{old}}), \nonumber \\
                        &\mathrm{s.t.} \sum_{i=1}^{ |\Ss| }q_i = 1, q_i \geq 0, \forall s.
\end{align}
is
\begin{equation}\label{eq:update-initial-state-prob}
  q_i = \frac{1}{R} \sum_{r=1}^{R} p(\smtr{s}{1}{r}=i | \ubmr{x}{r}; \bm{H}^{\mathrm{old}}), \forall\; i = 1, 2, \cdots, |\Ss|.
\end{equation}

\subsubsection{Transition Probability Update}
\eqref{eq:transition-update} can be written as
\begin{align}
  &\Qq(\bm{A}; \bm{H}^{\mathrm{old}})\nonumber \\
  % &= \sum_{r=1}^{R} \EE_{p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \left[\log\,\sum_{t=1}^{T^{(r)}-1}p(\smtr{s}{t+1}{r}|\smtr{s}{t}{r}; {A})\right] \nonumber\\
  &= \sum_{r=1}^{R} \sum_{\ubmr{s}{r}}{p(\ubmr{s}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \sum_{t=1}^{T^{(r)}-1}\log\,p(\smtr{s}{t+1}{r}|\smtr{s}{t}{r}; {A}) \nonumber \\
  &= \sum_{r=1}^{R} \sum_{t=1}^{T^{(r)}-1} \sum_{\smtr{s}{t}{r}=1}^{|\Ss|}\sum_{\smtr{s}{t+1}{r}=1}^{|\Ss|}{p(\smtr{s}{t}{r}, \smtr{s}{t+1}{r}| \ubmr{x}{r}; \bm{H}^{\mathrm{old}})} \log\,p(\smtr{s}{t+1}{r}|\smtr{s}{t}{r}; {A})
\end{align}
